\documentclass{article} \usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{minted}
\usepackage{ulem}


\title{eBPF Arena: A Tutorial}
\author{Farbod Shahinfar}
\begin{document}

\maketitle
{
    \small \textcolor{gray}{Estimated reading time: 10 minutes}
}

\include{inc.tex}

As part of my research, I spend time learning about eBPF.
Recently, I have been busy exploring \emph{Arena} --- a new eBPF API that
enables programs to allocate memory pages; similar functionality as \myC{mmap}
and \myC{munmap}~\cite{arenapatch}.

The eBPF community is doing a great job documenting the system and writing
tutorials. I felt I could contribute to this effort by writing about Arena.

%{
%    \textcolor{red}{At the moment, this blog post is a \emph{work in
%    progress} and I will update it as I learn more about Arena and figure
%    things out.}
%}

\section{Introduction}

The \emph{Arena} is a new MAP type (\bpfmap{arena}) available to eBPF programs
since kernel version 6.9.
This map is semantically different than previous ones.
Unlike data-structure MAPs (e.g., hash, array, bloom, stack, \dots{}) Arena
provides a direct access to the kernel memory instead of abstracting it away.
It increases the programs expressivity and enables them to implement their own
data-structures on-demand (e.g, a specific type of tree).

Three use-cases have been named for the Arena in the introduction
message~\cite{arenapatch} sent along its patch to the Linux mailing list. I
summarize them below:
\begin{itemize}
    \item User-space can use Arena's memory to implement its own
        data-structures (it is a normal memory region), while the memory is
        visible to eBPF programs which can implement a fast-path for operations
        like packet processing.
    \item Arena can be used as a communication channel between an eBPF and a
        user-space program.
    \item Arena can be used as a \emph{heap} of memory for the eBPF program
\end{itemize}
My understanding from these use-cases is that Arena provides access to raw
pages of memory shareable among eBPF programs and user-space. Each can decide
to use it as they see fit with their own responsibility. It is unlike other
types of data-structure MAPs that enforce certain constraints such as fixed key
and value sizes, or limiting the operations to lookup and update.

In this tutorial, I explore Arena's API, and provide some examples using it. I
hope this document helps others explore Arena and facilitate its development.
The example codes in this text are shared with you at
\href{https://github.com/bpf-endeavor/ebpf-arena-tutorial}{this GitHub  repository}.
I've only shared excerpts from the codes --- The parts I thought was important.
You may find and read the full examples in the repository.


\section{Arena's API}

\subsection{Declare a MAP}
A program can declare the Arena MAP similar to other MAPs. See Listing~\ref{lst:map_ex}
for an example. Here are some notes to keep in mind. The key size and value
sizes must be zero. The ``\myC{max_entries}'' field defines the maximum number
of pages for the map, and must be non-zero. As any other map, the maximum size
is limited to 4~GB. At the time of writing this text, Arena map declaration
supports three flags~\cite{arena_source}:
\begin{itemize}
    \item \myC{BPF_F_MMAPABLE}: \textbf{Always required}. Indicates the map should
        support memory mapping.
    \item \myC{BPF_F_SEGV_ON_FAULT}: The user-space program will receive a
        \myC{SEGFAULT} signal if the memory is not mapped by the eBPF program
        (e.g., out of range access on a list created by the eBPF program).
    \item \myC{BPF_F_NO_USER_CONV}: The memory region is private to the eBPF
        program and user-space program will not access it. This allows the JIT
        to perform optimizations avoiding address space casting (more about it
        later).
\end{itemize}
\textcolor{gray}{\textit{note: The \myC{BPF_F_MMAPABLE} must always be present.}}

\centering
\begin{listing}
\begin{minted}{c}
struct {
    __uint(type, BPF_MAP_TYPE_ARENA);
    __uint(map_flags, BPF_F_MMAPABLE);
    __uint(max_entries, 2); /* number of pages */
} arena SEC(".maps");
\end{minted}
    \caption{Example of using Arena map.}
    \label{lst:map_ex}
\end{listing}

\subsection{Helper functions}
Normal eBPF MAP helpers such as ``\myC{bpf_map_lookup_elem}'' are not
defined for Arena~\cite{arena_source}. Instead the following pair of functions
are available:
\begin{itemize}
    \item \myC{bpf_arena_alloc_pages}: Allocates a memory page
    \item \myC{bpf_arena_free_pages}: Frees the memory page
\end{itemize}

These functions are defined using kfunc~\cite{eunomia_kfunc, ebpf_docs_kfunc}.
As required by kfunc subsystem, you should declare the signature of these
helpers in your eBPF program to use these functions.
\begin{listing}
\begin{minted}{c}
void __arena* bpf_arena_alloc_pages(void *map, void __arena *addr,
    __u32 page_cnt, int node_id, __u64 flags) __ksym;
void bpf_arena_free_pages(void *map, void __arena *ptr,
    __u32 page_cnt) __ksym;
\end{minted}
\caption{Functions operating on Arena MAP.}
\label{lst:arena_kfuncs}
\end{listing}

Calling these functions may put the thread to sleep. For this reason the
functions are marked with \myC{KF_SLEEPABLE}~\cite{arena_source}, and the
verifier only allows sleepable eBPF programs to use them. An eBPF program has
sleeping privilege if the \myC{BPF_F_SLEEPABLE} flag was set when loading it
(see Listing~\ref{lst:loader} for an example).
Not all eBPF hooks support this flag. A list of sleepable
hooks are provided in \emph{libbpf} documentations~\cite{libbpf_sleepable}.

\textcolor{gray}{\emph{note: The memory pages allocated use the following flags:
\myC{GFP_KERNEL}, \myC{__GFP_ZERO}, \myC{__GFP_ACCOUNT}.}}

\subsection{Managing two address spaces}

\emph{Clang} compiler, the verifier, the JIT, and the runtime all work together
to make sure the eBPF program accesses the correct memory address. When using
Arena, there is a translation between the user address space and kernel address
space (unless the map is declared with ``\myC{BPF_F_NO_USER_CONV}'' flag
present). The eBPF program must mark the pointers with
``\myC{__attribute__((address_space(1)))}'' to let the Clang know about it and
cause the generation of ``\myC{bpf_arena_cast_user/kern}'' instructions. The
``\myC{__arena}'' tag used in the examples (and the accompanying repository),
especially when declaring variables and parameters, serves this purpose.

% \textcolor{gray}{\emph{note: if this explanation is vague, it is because I need to learn more about it.}}


\section{Examples}

I try to demonstrate how these APIs can be used to implement the use-cases
named above.

\subsection{Memory shared among eBPF and user-space}

As a first step, let us demonstrate how to share a counter between an eBPF
and an user-space program. For the sake of example, I used an eBPF program of type
\myC{BPF_PROT_TYPE_SYSCALL}~\cite{ebpf_docs_prog_syscall} which does not need to
be attached to a hook and can simply be invoked using a system call (for more
information search for \myC{BPF_PROG_TEST_RUN}~\cite{ebpf_docs_bpf_prog_run}).

Listing~\ref{lst:mogu} shows the code for the example program. In this
program, a global variable (``\myC{flag_initialized}'') tracks whether it is the first
time the program is executed or not. In the first run, a page is allocated
(using ``\myC{bpf_arena_alloc_pages}'') and its address is kept in
``\myC{mem}'' global variable. The allocated memory will hold a counter keeping
the number of invocations.

\textcolor{gray}{\textit{note: The global variable support is available since
kernel version 5.2~\cite{glb_var_post}.}}

\begin{listing}
\begin{minted}{c}
/* Declaration of Arena MAP as in Listing 1 ... */
static bool flag_initialized = false;
__arena void *mem = NULL;
SEC("syscall")
int mogu(void *_)
{
    bpf_printk("mogu: hello\n");
    if (!flag_initialized) {
        mem = bpf_arena_alloc_pages(&arena, NULL, 1, NUMA_NO_NODE, 0);
        if (mem == NULL) {
            bpf_printk("Failed to allocate memory");
            return 1;
        }
        flag_initialized = true;
    }
    if (mem == NULL) {
        /* this branch must never happen! */
        return 1;
    }
    __arena entry_t *e = mem;
    e->counter += 1;
    bpf_printk("counter: %lld\n", e->counter);
    return 0;
}
\end{minted}
\caption{An eBPF program using Arena.}
\label{lst:mogu}
\end{listing}

Next, in Listing~\ref{lst:loader}, we see how user-space program loads the eBPF
program and Listing~\ref{lst:loader_read} shows it accessing the counter on the
Arena memory page (The user-space relies on \emph{libbpf} and its skeleton
framework~\cite{libbpf_skeleton}).

\begin{listing}
\begin{minted}{c}
/* Some global vars */
static volatile int running = 0;
static int ebpf_prog_fd = -1;
static struct mogu *skel = NULL;

int main(int argc, char *argv[])
{
skel = mogu__open();
if (!skel) {
    fprintf(stderr, "Failed to open the skeleton\n");
    return EXIT_FAILURE;
}
/* Set sleepable flag */
bpf_program__set_flags(skel->progs.mogu, BPF_F_SLEEPABLE);
if (mogu__load(skel)) {
    fprintf(stderr, "Failed to load eBPF program\n");
    return EXIT_FAILURE;
}

ebpf_prog_fd = bpf_program__fd(skel->progs.mogu);
/* It will invoke the eBPF program for the first time */
handle_invoke_signal(0);

/* Keep running and handle signals */
running = 1;
signal(SIGINT, handle_signal);
signal(SIGHUP, handle_signal);
signal(SIGUSR1, handle_invoke_signal);
printf("Hit Ctrl+C to terminate ...\n");
printf("Invoke eBPF program:\n");
printf("\tMogu: pkill -SIGUSR1 mogu_loader\n");

while (running) { pause(); }

mogu__detach(skel);
mogu__destroy(skel);
printf("Done!\n");
return 0;
}
\end{minted}
\caption{User space program loading the program}
\label{lst:loader}
\end{listing}

\begin{listing}
\begin{minted}{c}
entry_t *e = skel->bss->mem;
if (e == NULL) {
    printf("NOTE: the initialization was not successful!\n");
    return;
}
printf("user: counter=%lld\n", e->counter);
\end{minted}
\caption{User-space accessing the memory page allocated from Arena}
\label{lst:loader_read}
\end{listing}


\subsection{Using Arena in XDP (or other hooks that can not sleep)}

As was mentioned, Arena memory allocation helper functions may put the calling
thread to sleep. This is not acceptable for programs attached to hooks such as
\emph{XDP}. But, this limitation does not mean that XDP programs can not access
the Arena memory. They just can not allocate or free the pages.

For the next step, Let us share the memory page allocated by the program
from the previous example with another eBPF program. This allows the second
program to use Arena backed memory pages without allocating memory itself.
Listing~\ref{lst:aloe} shows the second eBPF program.

Here are some details that I like to bring to your attention:
\begin{enumerate}
    \item Notice that both eBPF programs declare an Arena (as shown in
        Listing~\ref{lst:map_ex}). The loader (will be discussed next) will
        make sure both are using the same MAP using the libbpf's
        \myC{bpf_map__reuse_fd} helper.
    \item The second program expects the loader to pass the address of
        allocated page through its global variable called ``mem''. This address
        could have been shared among the two program in other ways, e.g.
        through a shared array MAP\@.
    \item The second program is using a non-standard helper named
        ``\myC{my_kfunc_reg_arena}''. This function (a kfunc defined by me)
        does not perform any operation. Its sole role is to let the verifier
        recognize that the program is using the Arena. See
        \S\ref{sec:need_custom} for more detail.
\end{enumerate}

\begin{listing}
\begin{minted}{c}
/* Declaration of Arena MAP as in Listing 1 ... */
__arena void *mem = NULL;

/* Load the kernel module for adding this kfunc */
long my_kfunc_reg_arena(void *p__map) __ksym;

SEC("syscall")
int aloe_main(void *_)
{
    my_kfunc_reg_arena(&arena_map);
    if (mem == NULL) {
        bpf_printk("aloe: not seeing the memory!\n");
        return 1;
    }
    __arena entry_t *e = mem;
    bpf_printk("aloe: counter=%lld\n", e->counter);
    e->counter += 100;
    return 0;
}
\end{minted}
\caption{An eBPF program that uses Arena pages allocated from another program}
\label{lst:aloe}
\end{listing}

To share the Arena from the first eBPF program (Listing~\ref{lst:mogu})
with the second one (Listing~\ref{lst:aloe}), the loader (user-space) program
has to mediate and pass the reference from first program to the second.
Listing~\ref{lst:share-map-ref} shows an example of this procedure.

\begin{listing}
\begin{minted}{c}
int arena_fd = bpf_map__fd(skel1->maps.arena);
bpf_map__reuse_fd(skel2->maps.arena_map, arena_fd);
/* Pass the pointer to the  second program */
skel2->bss->mem = skel1->bss->mem;
\end{minted}
\caption{Loader program assigning the Arena from the first program to the second program}
\label{lst:share-map-ref}
\end{listing}

\subsubsection{Why do we need a custom Arena function?}
\label{sec:need_custom}
If we remove the call to the non-standard \myC{my_kfunc_reg_arena} in
Listing~\ref{lst:aloe}, although the program would compile, the verifier will
complain that the program does not have any associated Arena map, but is trying
to use Arena memory addressing. The error message from verifier is as follows:

\begin{listing}
\begin{minted}{c}
addr_space_cast insn can only be used in a program that has an associated arena
\end{minted}
\caption{Verifier error message when the Arena is not referenced in the program}
\label{lst:verifier_err_msg}
\end{listing}

Looking at the verifiers source code to decipher the error message, we see that
the error is raised when the program is not associated with an Arena while the
program has an address space cast instruction~\cite{verifier_arena_not_set}
(exactly as written in the error message).

The association of an Arena MAP to a program happens when the it is referenced
(e.g., by using a helper function). Unfortunately, in the second program we
deliberately do not want to use Arena specific helpers. Other helper functions
such as \myC{bpf_map_lookup_elem} also cause other complains from the verifier
(in my test it complained due to attempting a zero byte read). Defining a new
helper that does nothing but still satisfies the verifier is a smart hack
around the situation. It showcases the flexibility that kfunc has brought to
eBPF ecosystem.

\textcolor{gray}{\textit{note: The idea of having a dummy helper was suggested
by my lab-mate \href{https://github.com/marcomole00/}{Marco Mole}.}}


\subsection{User-space managed data-structures}

In the third example, I demonstrate how user-space program can allocate pages
of Arena map (useful to implement its own custom data-structure). Since these
data-structures are on Arena pages, they will be accessible to eBPF programs
sharing the same Arena.

We can use \emph{libbpf} to get access to the Arena memory.
The library has an API named \myC{bpf_map__initial_value} (not well documented
at the moment). As input, this function receive a pointer to a MAP and a pointer
to a 64-bit variable. The return value is the address for the beginning of Arena
memory region (mapped to the user-space program memory address), also the
64-bit variable will be set to the size of the currently allocated
memory~\cite{libbpf_initial_value}.

To allocate memory pages, the user-space can cause page-faults by accessing
memory addresses in Arena. The page fault will notify the kernel to allocate
the memory page. For example, in our previous examples the Arena has two pages
(see Listing~\ref{lst:map_ex}). Assuming size of each page is 4 KB, the program
can access the Arena memory (the address we found using
\myC{bpf_map__initial_value}) at offset 0 and 4096 to allocate the two pages.
Accessing the memory outside of the Arena region will cause a \myC{SIGSEGV}
(Address boundary error).

The user-space program can use these memory pages to implement its own
data-structures in the Arena which is will be visible to eBPF programs.


\section{Final Notes}

The Arena feature is very new. It will probably change and improve. I am sure
by the time it is mature, eBPF community has written good documentation and
tutorials on it but meanwhile I hope this post fill the gap. It is important
for me to say that although I tried my best to share correct information, I can
not claim that I understand every aspect of Arena's implementation. There may
be misunderstandings on my side. In such a case please let me know and accept
my apology.

I want to conclude by listing some of the limitations of Arena that I observed
looking at the verifier and x86 JIT code.

\begin{enumerate}
  \item \sout{The atomic reads/writes are not supported in Arena memory region.} At the moment, in Arena memory region, limited atomic operations are support from eBPF side~\cite{arena_atomic}.
\item Arena pointer arithmetics are converted to 32 bit operations.
\item There can be only one Arena map per program of maximum size of 4 GB.
\item Program must be JIT compiled (it is almost always the case)
\item \myC{BPF_CAP} and \myC{CAP_PERFMON} is needed for loading a program with Arena
\item Sign extending load instruction is not supported for Arena memory region
\item The cast to address-space(1) will generate a ``\myC{dst = (u32)src}''
\item The cast lets the verifier know that the pointer is to arena
        the cast to address-space (0) turns the value into a scalar.
\item x86 JIT uses the R12 register to track the start of Arena virtual
    address (it was unused before).
\end{enumerate}

\subsection{Updates}
\begin{itemize}
  \item $1^{st}$ March, 2025:
    \begin{enumerate}
      \item Correction: There some \emph{atomic} instructions are available to eBPF programs to access Arena
      \item New information: There are on going work to support memory allocation in non-sleepable programs (e.g., XDP)~\cite{slub,try_alloc}. Thanks to \href{https://kkdwivedi.in/}{Kumar Kartikeya Dwivedi}
    \end{enumerate}
\end{itemize}

\bibliographystyle{plain}
\bibliography{ref}
\end{document}
